Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar

### 2018-12-04 accessPEP.Rmd
dist models!

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')

@@

## Preliminary data processing

### Cleaning data
```{r first cleaning}
## read in data
ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")

## percent with wrong dates by ctar

## percent with missing district by ctar

## exclude contacts

@@

@@

```{r doses}
ctar_metadata <- read.csv("data/ctar_metadata.csv")
mada_district <- readOGR("data/MadaGIS/MadaPops.shp")

img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
pal <- image_palette(img, 31)
ctar_metadata$color <- pal

ctar_metadata %>% 
  dplyr::select(id_ctar, starts_with("doses")) %>%
  gather(year, doses, -id_ctar) %>%
  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> doses

## getting dose time series by clinic
ctar_data %>%
  select(id_ctar, datej0, datej3, datej7, datej28) %>%
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
  gather(dose, date, -id_ctar) %>%
  group_by(id_ctar, date) %>%
  summarise(n = n()) -> dose_ts
dose_ts <- dose_ts[!is.na(dose_ts$date), ]
dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]

## getting theoretical # vials by year
dose_ts %>%
  mutate(year = year(date), vials = ceiling(n/2)) %>%
  group_by(id_ctar, year) %>%
  summarize(vials = sum(vials)) %>%
  left_join(doses) -> dose_comp

ggplot(data = dose_ts, aes(date, ctar)) + 
  geom_tile(aes(fill = n)) +
  scale_fill_continuous(type = "viridis", name = "Patients")

dose_ts %>%
  right_join(ts, by = c("date" = "value")) %>%
  select(id_ctar, date, n) %>%
  spread(id_ctar, n) %>%
  replace(., is.na(.), 0) -> doses_wide
dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])

date_mat <- replace(date_mat, is.na(date_mat), 1)
date_mat %>% 
  as_tibble() %>%
  group_by(date = year(doses_wide$date)) %>%
  summarise_all(funs(sum(.)/365)) -> clinic_reporting
colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date)
clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]

ggplot(clinic_reporting, aes(x = date, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")

@@

## districts that are non-reporter and data not available yet CTAR
no_data <- c("Manja", 'Ambatondrazaka', "Vangaindrano", "Tsiroanomandidy",
             "Antsiranana I", "Taolagnaro", "Ambovombe-Androy", "Nosy-Be", "Ambatomainty",
             "Marolambo", "Sainte Marie", "Antananarivo Renivohitra", "Antananarivo Avaradrano",
             "Ambohidratrimo", "Ankazobe", "Manjakandriana", "Anjozorobe", "Andramasina", 
             "Antananarivo Atsimondrano", "Fianarantsoa I", "Ambalavao", "Ambohimahasoa", 
             "Ikalamavony", "Lalangina", "Vohibato", "Isandra")
no_data <- mada_district$mdg_dis_co[match(no_data, mada_district$district)]
mada_district@data %>%
  select(district = mdg_dis_co, pop = pop2015adj, ttimes = study_area) -> covars

exposures %>%
  rename(date = year) %>%
  mutate(id_ctar = as.character(id_ctar)) %>%
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(!(district %in% no_data), prop >= 0.25) %>% # exclude no data & < 25% annual reporting
  group_by(district) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites)) %>% # average by ctar + district 
  left_join(covars) %>% # add pop + ttimes + reporting
  mutate(incidence = bites/(pop/1e5)) -> exps_cleaned 
exps_cleaned %$% plot(ttimes, bites)
exps_cleaned %$% plot(ttimes, incidence, bty = 
                        "n", pch = 20)

@@

ggplot() +
  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = incidence)) +
  scale_fill_continuous(type = "viridis") +
  geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
ggplot() +
  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = ttimes)) +
  scale_fill_continuous(type = "viridis") +
  geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))

library(lme4)
summary(lm(log(incidence) ~ ttimes, data = exps_cleaned))
```

>>>>

@@

## Preliminary data processing
```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')

@@

### Data
```{r get data}
## read in data
ctar_data <- read.csv("data/SaisieRage_DATA_2018-09-21_1755.csv")
ctar_metadata <- read.csv("data/ctar_metadata.csv")
mada_district <- readOGR("data/MadaGIS/MadaPops.shp")
dist_catchments <- read.csv("output/catchments_district_unmasked.csv", row.names = 1)
names(dist_catchments) <- c("mdg_dis_co", "CTAR", "ttimes_weighted")

## Centers with no data
no_data <- c("IPM", "Fianarantsoa", "Ambatomainty", "Ambovombe Androy",
             "Antsiranana", "Mandritsara", "Marolambo", "Nosy be", "Sainte Marie",
             "Taolagnaro", "Tsiroanomandidy", "Vangaindrano")
ctar_metadata$exclude <- 0
ctar_metadata$exclude[ctar_metadata$CTAR %in% no_data] <- 1

## Get color palette from other network fig!
img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
pal <- image_palette(img, 31)
add.alpha <- function(col, alpha=1){
  if(missing(col))
    stop("Please provide a vector of colours.")
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                       rgb(x[1], x[2], x[3], alpha=alpha))  
}
ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
ctar_metadata$fill <- ctar_metadata$color
ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)
ctar_metadata$ctar_dist <- mada_district$mdg_dis_co[match(ctar_metadata$District, 
                                                           mada_district$district)]
display_palette(ctar_metadata$color)
dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
mada_district@data <- left_join(mada_district@data, dist_catchments)
```

```{r catchment figure}
## get exposure matrix
ctar_data %>% 
  group_by(district, id_ctar) %>% # group by district and ctar
  summarize(n = n()) -> exp_mat # get count of exposures
dist_points <- cbind(select(mada_district@data,
                            mdg_dis_co), coordinates(mada_district))
names(dist_points)[2:3] <- c("dist_long", "dist_lat")
dist_points %>%
 left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
           by = c("mdg_dis_co" = "ctar_dist")) %>%
 mutate(dist_long = coalesce(dist_long, LONGITUDE), 
        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points

exp_mat %>% 
  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
            by = c("district" = "mdg_dis_co")) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines$size <- log(dist_lines$n + 0.1)*0.2

ctar_data %>%
  group_by(district) %>%
  summarize(n = n()) %>%
  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
             by = c("district" = "mdg_dis_co")) %>%
  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
dist_points$size <- log(dist_points$n + 0.1)*0.25

pdf("figs/network.pdf")
plot(mada_district, col = mada_district$fill, border = "grey") 
points(dist_points$dist_long, dist_points$dist_lat, pch = 20, cex = dist_points$size, 
       col = dist_points$color)
segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat, 
         lwd = dist_lines$size*2, col = dist_lines$color)
dev.off()

legend("topleft", c(">100", ">200", ">400", ">800"), title = "Number of total reported bites", 
       text.width = 3, pt.cex = circle_size, col = "grey",  pch = 16, inset = c(0.1, 0.1), lwd = 1.5, bty = "n", xpd = TRUE, y.intersp=2)

exp_mat %>% 
  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
  group_by(district) %>%
  mutate(prop = n/sum(n),
         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
ggplot(data = catch_check, aes(CTAR.x, district)) + 
  geom_tile(aes(fill = prop)) +
  scale_fill_continuous(type = "viridis", name = "proportion") +
  geom_text(label = catch_check$catch_match)
```

```{r}
## get vials delivered
ctar_metadata %>% 
  dplyr::select(id_ctar, starts_with("doses")) %>%
  gather(year, vials_delivered, -id_ctar) %>%
  mutate(year = as.numeric(substr(as.character(year), 7, 10))) -> vials_delivered

## getting dose time series by clinic
########### Need to fix dates first! ###############
ctar_data %>%
  select(id_ctar, datej0, datej3, datej7, datej28) %>%
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>%
  gather(dose, date, -id_ctar) %>%
  group_by(id_ctar, date) %>%
  summarise(n = n()) -> dose_ts
dose_ts <- dose_ts[!is.na(dose_ts$date), ]
dose_ts$ctar <- ctar_metadata$CTAR[match(dose_ts$id_ctar, ctar_metadata$id_ctar)]
dose_ts <- dose_ts[!is.na(dose_ts$ctar), ]

## getting theoretical # vials by year
dose_ts %>%
  mutate(year = year(date), vials = ceiling(n/2)) %>%
  group_by(id_ctar, year) %>%
  summarize(vials_estimated = sum(vials)) %>%
  left_join(vials_delivered) -> dose_comp

######### Patient time series
ctar_data %>%
  select(id_ctar, date_de_consultation) %>%
  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
  gather(dose, date_de_consultation, -id_ctar) %>%
  group_by(id_ctar, date_de_consultation) %>%
  summarise(n = n()) -> patient_ts
patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]

ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
  geom_tile(aes(fill = n)) +
  scale_fill_continuous(type = "viridis", name = "Patients")

patient_ts %>%
  right_join(ts, by = c("date_de_consultation" = "value")) %>%
  select(id_ctar, date_de_consultation, n) %>%
  spread(id_ctar, n) %>%
  replace(., is.na(.), 0) -> doses_wide
dose_mat <- as.matrix(doses_wide[, 2:(ncol(doses_wide) - 1)])

date_mat <- replace(date_mat, is.na(date_mat), 1)
date_mat %>% 
  as_tibble() %>%
  group_by(date = year(doses_wide$date_de_consultation)) %>%
  summarise_all(funs(sum(.)/n())) -> clinic_reporting
colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]
clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)

ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")
```

@@

```{r clean data}
## districts that are non-reporter and data not available yet CTAR
mada_district@data %>%
select(district = mdg_dis_co, pop = pop2015adj, ttimes_weighted, CTAR, exclude) -> covars

@@

exposures %>%
  rename(date = year) %>%
  left_join(covars) %>% # add pop + ttimes + reporting
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(exclude == 0, prop > 0.25) %>%
  group_by(district, date) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) -> exps_cleaned 
exps_cleaned %>%
  group_by(district) %>%
  summarise(bites = mean(bites)) %>%
  left_join(covars) -> exps_cleaned

library(lme4)

## poisson with total pop as offset
mod <- glm(round(bites) ~ ttimes_weighted + pop + offset(log(round(pop))), 
                 data = exps_cleaned, family = poisson)
plot(exps_cleaned$bites, predict(mod, type = "response"))
plot(exps_cleaned$ttimes_weighted, predict(mod, type = "response"))

```{r}
p_rabid = 0.50
mean_bites <- predict(mod, type = "response", newdata = covars)
bites <- rpois(length(mean_bites), mean_bites)
mean_exps <- runif(length(covars$pop), min = 0.39*0.01*exps_cleaned$pop/25, 
                    max = 0.39*0.01*exps_cleaned$pop/5)
rabid_exps <- rpois(length(mean_bites), mean_exps)
p_rabid_t <- ifelse(bites*p_rabid/rabid_exps > 0.95, (0.95*rabid_exps)/bites, p_rabid)
deaths <- rbinom(length(mean_bites), rabid_exps - round(bites*p_rabid_t), 0.16)
rho <- bites*p_rabid_t/rabid_exps
plot(covars$ttimes_weighted, rho)

```{r}
exps_cleaned %$% plot(ttimes_weighted, bites)

exps_cleaned %>%
  filter(exclude == 0) %$% 
  plot(ttimes_weighted, incidence, bty = 
                        "n", pch = 20)

@@

quartz()
ggplot() +
  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = CTAR)) +
  scale_colour_viridis_d()

geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
ggplot() +
  geom_polygon(data = mada_dist_gg, aes(x = long, y = lat, group = group, fill = ttimes_weighted)) +
  scale_fill_continuous(type = "viridis") +
  geom_point(data = ctar_metadata, aes (x = LONGITUDE, y = LATITUDE))
